### 2018-05-30 1_intro.tex
Final

As an example, consider two x86 assembly programs in Figure~\ref{simple_wmm_x86} (such little code examples that explain specific behaviour of the concurrent execution environment are called \textit{litmus tests}).
Two programs are the same, except the right-hand program uses the synchronisation instruction \texttt{MFENCE} after the store to the shared memory.
In both programs, the process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes to the~\texttt{y} and reads the~\texttt{x}.

\begin{figure}[t]
\centering
\ttfamily
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}[b]{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
& \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: allowed} \tabularnewline
\hline
\end{tabular}
\caption{Without synchronisation}
\label{simple_wmm_x86:nosync}
\end{subfigure}
\hfill
%
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MFENCE & MFENCE \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: forbidden} \tabularnewline
\hline
\end{tabular}
\caption{With synchronisation}
\label{simple_wmm_x86:sync}
\end{subfigure}
%
\caption{Store buffering~(SB): A litmus test illustrating the write-read reordering allowed by the \texttt{x86-TSO} memory model}
\label{simple_wmm_x86}
\end{figure}

Taking into account all possible interleavings of the read and write instructions, both litmus tests allow the following final states:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(0:EAX=0,~1:EBX=1)},
\item \texttt{(0:EAX=1,~1:EBX=0)},
\item \texttt{(0:EAX=1,~1:EBX=1)}.
\end{itemize}
%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' to be unreachable.
However, in the left-hand (non-synchronised) litmus test, the state `\texttt{(0:EAX=0,~1:EAX=0)}' is also allowed on the x86 architecture as its processors may cache the writes to the shared memory into their local \textit{store buffers} so that the updated value does not immediately become visible by processes running on other cores.
This behaviour is known as \textit{Store Buffering~(SB)}.
For preventing such a behaviour, the x86 architecture offers the synchronisation instruction \texttt{MFENCE} that flushes the store buffer of the process.
In Figure~\ref{simple_wmm_x86:sync}, this instruction is inserted for both processes after the store instructions, thus once it is executed in one process, the other process will read the updated value of this shared memory location.
The work~\cite{mckenney2010memory} gives a comprehensive review of common hardware architectures from the perspective of the structure of their caches and write buffers, that determine the effect of memory operations, and memory barriers they provide.

@@

In addition to developing the theoretical basis, researchers work on extracting the memory models for hardware architectures from existing implementations or from the specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade, memory models have been defined for most mainstream multiprocessor architectures, such as \texttt{x86-TSO} and \texttt{Sparc-TSO} (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.,~\cite{lustig2014pipecheck,lustig2016coatcheck}.

>>>>

As an example, consider two x86 assembly programs in Figure~\ref{simple_wmm_x86} (such little code examples that explain specific behaviour of the concurrent execution environment are called \textit{litmus tests}).
Two programs are the same, except the right-hand side program uses the synchronisation instruction \texttt{MFENCE} after the store to the shared memory.
In both programs, the process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes to the~\texttt{y} and reads the~\texttt{x}.

\begin{figure}[t]
\centering
\ttfamily
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}[b]{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
& \\
\hline
\multicolumn{2}{|l|}{exists (P0:EAX=0,\,P1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: allowed} \tabularnewline
\hline
\end{tabular}
\caption{Without synchronisation}
\label{simple_wmm_x86:nosync}
\end{subfigure}
\hfill
%
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MFENCE & MFENCE \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
\multicolumn{2}{|l|}{exists (P0:EAX=0,\,P1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: forbidden} \tabularnewline
\hline
\end{tabular}
\caption{With synchronisation}
\label{simple_wmm_x86:sync}
\end{subfigure}
%
\caption{Store buffering~(SB): A litmus test illustrating the write-read reordering allowed by the x86-TSO memory model}
\label{simple_wmm_x86}
\end{figure}

Taking into account all possible interleavings of the read and write instructions, both litmus tests allow the following final states:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(P0:EAX=0,~P1:EBX=1)},
\item \texttt{(P0:EAX=1,~P1:EBX=0)},
\item \texttt{(P0:EAX=1,~P1:EBX=1)}.
\end{itemize}
%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(P0:EAX=0~/\textbackslash~P1:EAX=0)}' to be unreachable.
However, the non-synchronised litmus test in Figure~\ref{simple_wmm_x86:nosync} allows the state `\texttt{(P0:EAX=0,~P1:EAX=0)}' on the x86 architecture as x86 processors may cache the writes to the shared memory into their local \textit{store buffers} so that the updated value does not immediately become visible by processes running on other cores.
This behaviour is known as \textit{Store Buffering~(SB)}.
For preventing such a behaviour, the x86 architecture offers the synchronisation instruction \texttt{MFENCE} that flushes the store buffer of the process.
In Figure~\ref{simple_wmm_x86:sync}, this instruction is inserted for both processes after the store instructions, thus once it is executed in one process, the other process will read the updated value of this shared memory location.
The work~\cite{mckenney2010memory} gives a comprehensive review of common hardware architectures from the perspective of the structure of their caches and write buffers, that determine the effect of memory operations, and memory barriers they provide.

@@

In addition to developing the theoretical basis, researchers work on extracting the memory models for hardware architectures from existing implementations or from the specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade, memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.,~\cite{lustig2014pipecheck,lustig2016coatcheck}.

### 2018-05-30 2_wmm.tex
Final

The set of memory events $\mathbb{M}$ is divided into write events $\mathbb{W}$ (that write values to shared-memory locations) and read events $\mathbb{R}$ (that read values stored in shared-memory locations).
We add a restriction that each memory event uses at most one shared location, so that the write instruction $i = write(l_1, l_2)$, that encodes the write from the shared location $l_2$ to the shared location $l_1$, is represented as two consequent events $e_1~=~\texttt{load}(r_1~\leftarrow~l_2); \ e_2~=~\texttt{store}(l_1~\leftarrow~r_1)$.
Also, it is important to separate the set of initial write events $\mathbb{IW}~\subseteq~\mathbb{W}$ that perform initialisation of program variables.

@@

\begin{figure}[!htb]
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(0:EAX=0,~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Candidate executions for the litmus test in Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

>>>>

The set of memory events $\mathbb{M}$ is divided into write events $\mathbb{W}$ (that write values to shared-memory locations) and read events $\mathbb{R}$ (that read values stored in shared-memory locations).
We add a restriction that each memory event uses at most one shared location, so that the write instruction $i = write(l_1, l_2)$, that encodes the write from the shared location $l_2$ to the shared location $l_1$, is represented as two consequent events $e_1\,=\,{\texttt{load}(r_1\,\leftarrow\,l_2)}; \ e_2\,=\,{\texttt{store}(l_1\,\leftarrow\,r_1)}$.
Also, it is important to separate the set of initial write events $\mathbb{IW}~\subseteq~\mathbb{W}$ that perform initialisation of program variables.

@@

\begin{figure}[!htb]
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(P0:EAX=1,~P1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(P0:EAX=1,~P1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(P0:EAX=1,~P1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(P0:EAX=0,~P1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Candidate executions for the litmus test in Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

### 2018-05-30 3_encod.tex
Cleanup finale up to Xevent.
Final

At present time, satisfiability solvers are behind the numerous verification and bounded model checking tools.
Although, as most modern programming languages are typed, the SAT-formula that encodes the constraints of a program must include non-boolean constraints encoded as boolean variables, which is ineffective, therefore most software analysis and verification tools use \textit{Satisfiability Modulo Theories~(SMT)} -solvers, that integrate first-order reasoning with the background theory reasoning~\cite{kroening2008decision}.
For example, the Integer theory includes operators `$+$', `$-$', `$\geqslant$' and others).
Most modern SMT-solvers support integer arithmetic, real arithmetic, fixed-size bit-vectors, arrays, and other theories.
The direction of the development of SMT-solvers is provided by the \texttt{SMT-LIB} standard~\cite{smt-lib}, which provides definition of the unified language for defining SMT-formulas, and description of most common underlying logics and theories.

@@

To encode the data-flow constraints, we use the \textit{Static Single-Assignment~(SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula.
The SSA form requires each variable to be assigned only once within entire program.
In contrast, \porthos{} used the \textit{Dynamic Single-Assignment~(DSA)} form, that requires indices to be unique only within a control-flow branch.
Although the number of variable references (each of which is encoded as unique SMT-variable) on average is logarithmically less in the case of the DSA form than the SSA form, the result SMT-formula still needs to be complemented by same number of equality assertions when encoding the data-flow of merge points~\cite{Porthos17a}.

>>>>

At present time, satisfiability solvers are behind the numerous verification and bounded model checking tools.
Although, as most modern programming languages are typed, a SAT-formula that encodes the constraints of a program must include non-boolean constraints encoded as boolean variables, which is ineffective, therefore most software analysis and verification tools use \textit{Satisfiability Modulo Theories~(SMT)} solvers, that integrate first-order reasoning with the background theory reasoning~\cite{kroening2008decision}.
For example, the Integer theory includes operators `$+$', `$-$', `$\geqslant$' and others).
Most modern SMT-solvers support integer arithmetic, real arithmetic, fixed-size bit-vectors, arrays, and other theories.
The direction of the development of SMT-solvers is provided by the \textit{SMT-LIB standard}~\cite{smt-lib}, which provides definition of the unified language for defining SMT-formulas, and description of most common underlying logics and theories.

@@

To encode the data-flow constraints, we use the \textit{Static Single-Assignment~(SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula.
The SSA form requires each variable to be assigned only once within entire program.
In contrast, \porthos[1] uses the \textit{Dynamic Single-Assignment~(DSA)} form, that requires indices to be unique only within a control-flow branch.
Although the number of variable references (each of which is encoded as unique SMT-variable) on average is logarithmically less in the case of the DSA form than the SSA form, the result SMT-formula still needs to be complemented by same number of equality assertions when encoding the data-flow of merge points~\cite{Porthos17a}.
